Extended Data

### 2022-05-09 doc/submissions/2022_04_Nature/Response to reviews.Rmd

**We agree with this comment and specify our motivation in the methods section: "We defined CTW for *DOY~25~*, as opposed to other parameters describing the timing of growth, because spring temperatures should have the most direct influence on this variable."**

**We also repeat the analyses using a definition of CTW that is independent of *DOY~25~* and all other response variables. As explained in the methods, "To ensure that patterns were robust under an alternative definition of CTW, and to parallel the monthly time windows used in our tree-ring analysis (detailed below; Fig. 3, Extended Data Figure 6-7), we also ran analyses where we fixed the CTW identified by climwin to be the month with the most days in the CTW (Extended Data Table 2) each critical window."**

I appreciated the clarifications done concerning the data cleaning process. However, I didn’t find in the manuscript the quantity of data that have been removed at it stages. I believe it is worth mentioning it.

**We simplified the presentation of our data cleaning methods, ensuring that these numbers are clearly presented for both stagges of cleaning:**

- **Paragraph describing screening of raw dendrometer band data: "In total, this process removed 309 of the 2708 available tree-year records for 2011-2020 at SCBI and 1999-2003 at Harvard Forest."**
- **Paragraph describing removal of poorly fit records: "In total, this process removed 189 of the 2399 tree-year records deemed appropriate for analysis, leaving a total of 2210 tree-years included in the final analysis (Extended Data Table 1)."**

### 2022-05-09 doc/manuscript/Dow_Extended_Data.Rmd
add map source
#137

@camerondow35 , FYI. Not sure if we should add any more info here.
clarify dendro band n's
#141

## table captions
species_table_caption <-"Extended Data Table 1 | Dominant ring- and diffuse-porous species at the Smithsonian Conservation Biology Institute (SCBI) and Harvard Forest, along with sample sizes included in this analysis."
mean_growth_parameters_caption = "Extended Data Table 2 | Summary of parameters describing the seasonality and temperature senisitivity of ring- and diffuse- porous species groups at the Smithsonian Conservation Biology Institute (SCBI) and Harvard Forest."
tree_ring_summary_caption =  "Extended Data Table 3 | Summary of tree-ring chronologies analyzed and number of significant (at p=0.05) positive or negative correlations of ring width index to monthly $T_{max}$ in univariate and multivariate analyses."

@@

![**Extended Data Figure 1 | Map of sampling locations of tree-ring chronologies analyzed in this study.** Sites are colored by the xylem porosity type of species sampled: ring poroous (RP), diffuse porous (DP), or both. Sampling details are provided in `r chron_table`.](tables_figures/chronology_map.png)

>>>>

## table captions
species_table_caption <-"Extended Data Table 1 | Dominant ring- and diffuse-porous species at the Smithsonian Conservation Biology Institute (SCBI) and Harvard Forest, along with sample sizes included in our final analysis."
mean_growth_parameters_caption = "Extended Data Table 2 | Summary of parameters describing the seasonality and temperature senisitivity of ring- and diffuse- porous species groups at the Smithsonian Conservation Biology Institute (SCBI) and Harvard Forest."
tree_ring_summary_caption =  "Extended Data Table 3 | Summary of tree-ring chronologies analyzed and number of significant (at p=0.05) positive or negative correlations of ring width index to monthly $T_{max}$ in univariate and multivariate analyses."

@@

![**Extended Data Figure 1 | Map of sampling locations of tree-ring chronologies analyzed in this study.** Sites are colored by the xylem porosity type of species sampled: ring poroous (RP), diffuse porous (DP), or both. Sampling details are provided in `r chron_table`. Base map source is ggplot2.](tables_figures/chronology_map.png)

### 2022-05-09 doc/manuscript/Dow_manuscript_Nature.Rmd
clarify dendro band n's
#141

Metal dendrometer bands were installed on 941 trees within the SCBI and Harvard Forest ForestGEO plots.
Bands were placed on dominant species, including two diffuse- and two ring porous species at SCBI and eight diffuse- and three ring porous species at Harvard Forest (`r EDT_species`).
While we do not estimate the ages of trees in our sample, bands at both sites were placed on individuals of differing sizes in an attempt to measure trees across a range of ages.
Bands were measured with a digital caliper approximately every 1-2 weeks within the growing season from 2011-2020 at SCBI and 1998-2003 at Harvard Forest. 
The number of bands measured at each site fluctuated slightly as trees were added or dropped from the census (e.g., because of tree mortality). 
Across years, the number of bands sampled averaged 129 (range: 91-138) at SCBI and 717 (range: 700-755) at Harvard Forest. <!--verified - these are the pre-cleaning numbers-->
In total, our analysis included `r n_tree_years` tree-years (`r  EDT_species`).

@@

The raw dendrometer band data were screened to remove records or entire tree-years that were inappropriate for our analysis because of reduced reliability of predicted growth in the modeled curves.
Specifically, we removed tree-years with small or negligible total growth ($\Delta DBH$ $\le 0.005$ cm; SCBI = 26, Harvard Forest = 253) and tree-years where the first intra-annual measurement was later than the first spring survey (trees that were missed in the initial census; SCBI = 22, Harvard Forest = 8).
In total, this process removed `r 26+253+22+8` of the `r n_tree_years` available records for 2011-2020 at SCBI and 1999-2003 at Harvard Forest.

@@

After fitting the growth model, we removed tree-years with poor fits.
Models were judged to be poorly fit if modeled growth parameters were outliers, which were commonly indicative of unrealistic fits (e.g., growth occurring outside the growing season or over a very short period) and underlain by very slow tree growth or poor data records that passed the initial screening (described above).
Modeled fits for tree-years were removed under two conditions:
(1) *g~max~* was $\ge 2.5$ standard deviations away from the mean for each site-xylem architecture group combination (SCBI = 3, Harvard Forest = 11);
(2) timing variables (*DOY~ip~*, *DOY~25~*, *DOY~50~*, *DOY~75~*) were $\ge 2.5$ standard deviations away from the means for their site, xylem architecture group, and year (SCBI = 74, Harvard Forest = 101).
In total, this process removed `r 3+11+74+101` of the `r n_tree_years - 26+253+22+8` records deemed appropriate for analysis.
At both sites the tree-years removed through this method were proportional to the original sample size, indicating that no species or size class was disproportionately removed compared to others.
This process was repeated using 2 and 3 standard deviations as the cutoff for defining outliers, yielding qualitatively similar results.

>>>>

Metal dendrometer bands were installed on 941 trees within the SCBI and Harvard Forest ForestGEO plots.
Bands were placed on dominant species, including two diffuse- and two ring porous species at SCBI and eight diffuse- and three ring porous species at Harvard Forest (`r EDT_species`).
While we do not estimate the ages of trees in our sample, bands at both sites were placed on individuals of differing sizes in an attempt to measure trees across a range of ages.
Bands were measured with a digital caliper approximately every 1-2 weeks within the growing season from 2011-2020 at SCBI and 1998-2003 at Harvard Forest. 
The number of bands measured at each site fluctuated slightly as trees were added or dropped from the census (e.g., because of tree mortality). 
Across years, the number of bands sampled averaged 129 (range: 91-138) at SCBI and 717 (range: 700-755) at Harvard Forest. <!--verified - these are the pre-cleaning numbers-->

@@

The raw dendrometer band data were screened to remove records or entire tree-years that were inappropriate for our analysis because of reduced reliability of predicted growth in the modeled curves.
Specifically, we removed tree-years with small or negligible total growth ($\Delta DBH$ $\le 0.005$ cm; SCBI = 26, Harvard Forest = 253) and tree-years where the first intra-annual measurement was later than the first spring survey (trees that were missed in the initial census; SCBI = 22, Harvard Forest = 8).
In total, this process removed `r 26+253+22+8` of the `r n_tree_years + (3+11+74+101) + (26+253+22+8)`<!--first set is n removed for poor model fit, 2nd set is n removed here--> available tree-year records for 2011-2020 at SCBI and 1999-2003 at Harvard Forest.

@@

After fitting the growth model, we removed tree-years with poor fits.
Models were judged to be poorly fit if modeled growth parameters were outliers, which were commonly indicative of unrealistic fits (e.g., growth occurring outside the growing season or over a very short period) and underlain by very slow tree growth or poor data records that passed the initial screening (described above).
Modeled fits for tree-years were removed under two conditions:
(1) *g~max~* was $\ge 2.5$ standard deviations away from the mean for each site-xylem architecture group combination (SCBI = 3, Harvard Forest = 11);
(2) timing variables (*DOY~ip~*, *DOY~25~*, *DOY~50~*, *DOY~75~*) were $\ge 2.5$ standard deviations away from the means for their site, xylem architecture group, and year (SCBI = 74, Harvard Forest = 101).
In total, this process removed `r 3+11+74+101` of the `r n_tree_years + (3+11+74+101)` tree-year records deemed appropriate for analysis, leaving a total of `r n_tree_years` tree-years included in the final analysis (`r  EDT_species`).
At both sites the tree-years removed through this method were proportional to the original sample size, indicating that no species or size class was disproportionately removed compared to others.
This process was repeated using 2 and 3 standard deviations as the cutoff for defining outliers, yielding qualitatively similar results.
